---
title: "Microbiome selection"
author: "Jigyasa Arora and Sasha Mikheyev"
date: "2/16/2018"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 5
    theme: cerulean
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache=TRUE, fig.path='figures/', fig.width=5, fig.height=5 )
```

# Phenotypic analysis

generating the correct format for all the fly eclosion time files-

```{r loadLibraries, message=FALSE}
library(knitr)
library(kableExtra)
library(gridExtra)
library(scales)
library(RColorBrewer)
library(tidyverse) #dplyr package included
library(stringr)
library(lme4)
library(nlme)
library(effects)
library(DESeq2)
library(qiime2R)
library(phyloseq)
library(DivNet)
library(vegan)
library(sjPlot) #  mixed effects table summary

theme_set(theme_minimal() + theme(legend.position = "bottom", axis.title =  element_text(size = 20), axis.text =  element_text(size = 10), strip.text = element_text(size = 20)))
```


```{r loadData}
flydata <- read_csv("data/flydata.csv") %>% mutate(id = paste(diet, selection, round, line, vial), time = emergence_time, replicate = paste(diet, selection, round, line))
```

## Mixed model analysis

```{r modRandom}
flydataFactors <- flydata %>% mutate(day = ifelse(time<225, 1, ifelse(time<240, 2, ifelse(time<270, 3, 4)))) %>% as.data.frame()
flydataFactors$selection <- factor(flydataFactors$selection)
flydataFactors$diet <- factor(flydataFactors$diet)
mod <- lmer(time ~ round * diet * selection + (1|line:replicate), data = flydata )
hist(resid(mod))

tab_model(mod, show.icc = FALSE, show.re.var = FALSE, pred.labels = c("Intercept (High-sugar diet, No-selection control)", "Round of selection", "No-sugar diet", "Selection treatment", "Round of selection × No-sugar diet", "Round of selection × Selection treatment", "No-sugar diet × Selection treatment", "Round of selection × No-sugar diet× Selection treatment"))

effect(term="round:selection", xlevels=list(selection=c("selection","noselection")), mod=mod) %>% as.data.frame() %>% ggplot(aes(round,fit,color=selection))+geom_line()

effect(term="round:diet", xlevels=list(diet=c("control","nsd")), mod=mod) %>% as.data.frame() %>% ggplot(aes(round,fit,color=diet))+geom_line()

effect(term="round:diet:selection", xlevels=list(selection=c("selection","noselection"),diet=c("control","nsd")), mod=mod) %>% as.data.frame()%>%
  ggplot(aes(round,fit,color=selection,linetype=diet))+geom_line()

diet_labeller <- function(variable, value) {
  diet_names <- list('nsd'='No-sugar diet', 'hsd'='High-sugar diet')
  selection_names <- list('selection'='Selection', 'noselection'='No Selection')
  if (variable == 'selection')
    return(selection_names[value])
  else if (variable == 'diet') 
    return(diet_names[value])
  else
    return(as.character(value))
}
effect(term="round:diet:selection", xlevels=list(selection=c("selection","noselection"),diet=c("control","nsd")), mod=mod) %>%
  as.data.frame() %>% 
  left_join(flydata, by = c("round", "diet", "selection")) %>% 
  ggplot(aes(round-1, fit, color=selection)) + geom_boxplot(aes(round-1, time, group = interaction(round,selection)), alpha = 0.1) +
  geom_line() + geom_ribbon(aes(ymin=lower,ymax=upper, linetype = NA), alpha=0.3) + 
  facet_grid(diet~., scales="free", labeller=diet_labeller) + 
  scale_color_manual(name = "Selection", labels = c("Random selection control", "Mean eclosion time"), values = c("dodgerblue", "orange")) + ylab("Eclosion time (hours)") + xlab("Round of selection") + guides(alpha = F, color=guide_legend(override.aes=list(fill=NA))) 

effect(xlevels=list(round=c(1,5)), term="round:diet",  mod=mod) %>% as.data.frame()

#ggsave("figures/modeFit.pdf", height=7, width =5)

ggplot(flydataFactors, aes(day, fill=selection))+geom_histogram(stat="count", position="dodge") +facet_grid(.~diet) 

tradeoff <- flydata %>% group_by(diet, round, selection, line, replicate) %>% summarize(time = mean(time), flies = n()) 
tradeoff %>% filter(round == 5) %>% ggplot(aes(time, flies, color = selection, shape = as.factor(round))) + geom_point() + facet_grid(.~diet, scales = "free") + stat_smooth(method="lm", aes(group = 1)) 
```

Selection has no effect on fly emergence, though emergence time decreases throughout the experiment in any case.

In round 4 there are no flies eclosing in the fourth day. Maybe some problem? Does this affect results?

```{r check4}
mod2 <- lme(time ~ round * diet * selection , random=~1|line/replicate, method="REML", data = flydata %>% filter(round != 4))
summary(mod2)
```

The model is robust to dropping this round entirely

# Microbial analysis

## Load data manually from qiime2

```{r microbial, message=F}
otus <- read_qza("data/table.qza")
taxonomy <- read_qza("data/taxonomy.qza")
tree<-read_qza("data/rooted-tree.qza")
taxonomy<-read_qza("data/taxonomy.qza")
tax_table <- data.frame(taxon = as.character(taxonomy$data$Taxon)) %>%  separate(taxon, c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species"), sep = ";") %>% mutate_all(funs(gsub(".*__", "", .))) 
rownames(tax_table) <- taxonomy$data$Feature.ID
tax_table <- as.matrix(tax_table)
# loading metadata
metadata <- read.table("data/metadata.txt", sep='\t', header=T, row.names = 1, comment="") 
   # create a bunch of category columns for divnet and append phenotypic data
metadata <- metadata %>% 
  mutate( sample_names = rownames(metadata), 
          round = round - 1,
          roundSelection = paste(round, selection), 
          roundSelectionDiet = paste(round, selection, diet),  
          id = paste(diet, selection, round, line, vial),
          line = factor(line)) %>%
  left_join(flydata %>% group_by(id) %>% summarize(time = mean(time))) 
rownames(metadata) <- metadata$sample_names 

# make final product
microbedat <- phyloseq(otu_table(otus$data, taxa_are_rows = T), phy_tree(tree$data), tax_table(tax_table), sample_data(metadata))
```

### Descriptive bar plots

These are not super-useful, but highlight the difference between diets

```{r alpha, message = FALSE, warning = FALSE, results = 'hide'}
abundantClasses <- psmelt(microbedat) %>% group_by(Genus) %>% summarize(frac = n() / nrow(.)) %>% filter(frac > 0.01) %>% na.omit() %>% pull(Genus)

psmelt(microbedat) %>% mutate(round = round ) %>%filter(Genus %in% abundantClasses) %>% ggplot(aes_string(x = "round", y = "Abundance", fill = "Genus")) + geom_bar(stat = "identity", position = "fill") + theme(axis.text.x = element_text(angle = 0, hjust = 0), legend.text=element_text(size=rel(1)), legend.title=element_blank()) + scale_y_sqrt() + ylab(expression(sqrt("Fraction of community"))) + xlab("Round of selection") + facet_grid(diet~selection, scales="free", labeller=diet_labeller) + scale_fill_brewer(palette = "Set3") 
ggsave("figures/barplot.pdf", height = 7, width = 7)
```

## Alpha diversity

```{r divnetAll, cache=T, message = FALSE, warning = FALSE, results = 'hide'}
divnet_all <-  divnet(tax_glom(microbedat, taxrank = "Genus"), X = "roundSelectionDiet", ncores = 20)

divnet_summary <- divnet_all$shannon %>% summary %>% left_join(metadata) %>% group_by(roundSelectionDiet) %>% summarize(alpha = estimate[1], lower = lower[1], upper = upper[1], round = round[1], selection = selection[1], diet = diet[1]) 
```


```{r divnetAllSummary}
divnet_summary %>% ggplot(aes(selection, alpha, color = selection)) + geom_point() + geom_linerange(aes(ymin = lower, ymax = upper)) + facet_grid(diet~round, scales = "free") + theme_minimal() + theme(axis.text.x=element_blank()) + xlab("Round of evolution") + guides(color = F)
ggsave("figures/divnet.pdf")
```

### Alpha diversity using a less sophisticated approach
```{r naiveRichness}
(p <- plot_richness(microbedat, "round", measures = "Shannon", color = "selection") + facet_grid(diet~., scales = "free_y", labeller=diet_labeller) + stat_smooth(method="lm")  + scale_color_manual(name = "Selection", labels = c("Random selection control", "Mean eclosion time"), values = c("dodgerblue", "orange")) + ylab("Shannon diversity") + xlab("Round of selection") + theme(axis.text.x = element_text(angle = 0, hjust = 0)) )

tab_model(lm(value ~ diet*round, data = p$data), pred.labels = c("Intercept (High-sugar diet)", "No-sugar diet", "Round of selection",  "Round of selection × No-sugar diet"))

ggsave("figures/shannon.pdf", height = 6, width = 6)
```

## Examining microbial community differentiation as in the mixed model

```{r adonis}
d = UniFrac(tax_glom(microbedat, taxrank = "Genus"))
d.mds <- metaMDS(d, zerodist=ignore)

scrs <- cbind(metadata, data.frame(MDS1 = d.mds$points[,1], MDS2 = d.mds$points[,2])) 

cent <- cbind(metadata, data.frame(MDS1 = d.mds$points[,1], MDS2 = d.mds$points[,2])) %>% aggregate(cbind(MDS1, MDS2) ~ diet + round + selection, data = ., FUN = mean) 

segs <- merge(scrs, setNames(cent, c('diet', 'round','selection' ,'oNMDS1','oNMDS2')),
              by = c('diet', 'round', 'selection'), sort = FALSE)

ggplot(scrs, aes(x = MDS1, y = MDS2, shape = selection, color = factor(round))) +
  facet_wrap(~diet) + scale_colour_brewer(palette = "YlOrRd") + 
  geom_segment(data = segs,
               mapping = aes(xend = oNMDS1, yend = oNMDS2)) + # spiders
  geom_point(data = cent, size = 3) +                         # centroids
  geom_point() +                                              # sample scores
  coord_fixed()                                               # same axis scaling

(microbes.adonis <- adonis2(d ~ diet * selection * round, as(sample_data(microbedat), "data.frame"), permutations = 10000, strata = line))

anova(betadisper(d, sample_data(microbedat)$diet))
plot(betadisper(d, sample_data(microbedat)$diet))
microbes.adonis
```

There are dispersion differences between the diets, so let's analyze them separately

### HSD permutation test

```{r}
microbedat.hsd <- tax_glom(subset_samples(microbedat, diet == "hsd"), taxrank = "Genus")  
d.hsd = UniFrac(microbedat.hsd)
(hsd.adonis <- adonis2(d.hsd ~  selection * round, as(sample_data(microbedat.hsd), "data.frame"), permutations = 10000, strata = line))
```

### NSD permutation test

```{r}
microbedat.nsd <- tax_glom(subset_samples(microbedat, diet == "nsd"), taxrank = "Genus")  
d.nsd = UniFrac(microbedat.nsd)
(hsd.adonis <- adonis2(d.nsd ~  selection * round, as(sample_data(microbedat.nsd), "data.frame"), permutations = 10000, strata = line))

anova(betadisper(d.nsd, sample_data(microbedat.nsd)$round))

```

Looks like there is evidence of community change in the low sugar diet over time, but no detectable change in the high sugar diet.

## Changes in OTU abundance

### As a function of fly emergence time

```{r exmergence_time}
microbedat.phyloseq.time <- phyloseq_to_deseq2(subset_samples(microbedat, diet == "hsd" & time > 0), ~ time + line)
microbedat.phyloseq.time <- results(DESeq(microbedat.phyloseq.time, test="LRT", fitType="local", reduced = ~ line))
head(microbedat.phyloseq.time[order(microbedat.phyloseq.time$pvalue),])

microbedat.phyloseq.time <- phyloseq_to_deseq2(subset_samples(microbedat, diet == "nsd" & time > 0), ~ time + line)
microbedat.phyloseq.time <- results(DESeq(microbedat.phyloseq.time, test="LRT", fitType="local", reduced = ~ line))
head(microbedat.phyloseq.time[order(microbedat.phyloseq.time$pvalue),])

```
No individual OTU is associated with fly emergence tim

### As a function of round

```{r phyloseq_round}
microbedat.phyloseq.round <- phyloseq_to_deseq2(tax_glom(subset_samples(microbedat, diet == "hsd"), taxrank = "Genus"), ~ round + line)
microbedat.phyloseq.round <- results(DESeq(microbedat.phyloseq.round, test="LRT", fitType="local", reduced = ~  line))
head(microbedat.phyloseq.round[order(microbedat.phyloseq.round$pvalue),])

microbedat.phyloseq.round <- phyloseq_to_deseq2(tax_glom(subset_samples(microbedat, diet == "nsd"), taxrank = "Genus"), ~ round + line)
microbedat.phyloseq.round <- results(DESeq(microbedat.phyloseq.round, test="LRT", fitType="local", reduced = ~  line))
head(microbedat.phyloseq.round[order(microbedat.phyloseq.round$pvalue),])
head(microbedat.phyloseq.round[order(microbedat.phyloseq.round$pvalue),])
```

# Session info
```{r sessioninfo, echo=F}
sessionInfo <- devtools::session_info() 
sessionInfo$platform 
sessionInfo$packages %>% kable("html") %>% kable_styling() %>% scroll_box(height = "400px") 
```